---
title: "Final_Project_MLM"
output: html_document
date: "2024-12-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Front Matter

```{r}
# Add libraries if needed
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(lme4)

# Read in dataset
TestScores <- read.csv(file = "/Users/hocaroline/Downloads/test_scores.csv")
```

```{r}
# Select relevant columns
TestScores <- 
  TestScores %>%
  select(teaching_method, lunch, posttest, classroom)

head(TestScores)
```

```{r}
#Find mean and standard deviation of the posttest scores 
Posttest_Score <- TestScores %>%
  summarize(overallmean = mean(posttest, na.rm = TRUE),
            overallsd = sd(posttest, na.rm = TRUE))
kable(Posttest_Score, format = "markdown", col.names = c("Overall Mean", "Overall Standard Deviation"), align = c("c", "c"))

# Histogram that describes the distribution of posttest scores for individual student
posttest_all <- ggplot(data = TestScores, mapping = aes(x = posttest)) +
  geom_histogram(binwidth = 2, fill = "white", color = "black") +
  labs(title = "Distribution of Post-test Scores for Individual Students", 
       x = "Post-test Scores",
       y = "Number of Students") +
  theme(
    plot.title = element_text(size = 15L,
    face = "italic",
    hjust = 0.5))
```

```{r}
#Find the mean and standard deviation for each type of classroom and form a histogram that represent the distribution of the mean posttest scores for different types of classrooms
posttest_agg <- TestScores %>%
  group_by(classroom) %>%
  summarize(mean_score = mean(posttest, na.rm = TRUE),
            sd_score = sd(posttest, na.rm = TRUE)) %>%
  ggplot(mapping = aes(x = mean_score)) +
     geom_histogram(binwidth = 2, fill = "white", color = "black") +
     labs(title = "Distribution of Mean Posttest Scores for Classrooms", 
          x = "Mean Posttest Scores",
          y = "Number of Classrooms") +
  theme(
    plot.title = element_text(size = 15L,
    face = "italic",
    hjust = 0.5))
```

```{r}
grid.arrange(posttest_all, posttest_agg, ncol = 1)
```

### Level one categorical variable (Lunch)
```{r}
# Find the counts of students for each lunch type
lunch_count <- table(TestScores$lunch)
kable(lunch_count, format = "markdown", col.names = c("Lunch Type", "Count"), align = c("c", "c"))

#Find the proportion of students for each lunch type
lunch_prop <- prop.table(table(TestScores$lunch))
kable(lunch_prop, format = "markdown", col.names = c("Lunch Type", "Proportion"), align = c("c", "c"))

#Use bar graph to compare the number of students who does or does not qualify for reduced/free lunch 
ggplot(data = TestScores, mapping = aes(x = lunch)) + 
  geom_bar(fill = "white", color = "black", width = 0.5) +
  labs(x = "Lunch Type",
       y = "Frequency",
       title = "Number of Students for Each Lunch Type") +
  theme(
    plot.title = element_text(size = 15L,
    face = "italic",
    hjust = 0.5),
    axis.text.y = element_text(size = 11L),
    axis.text.x = element_text(size = 11L),
    axis.title.y = element_text(size = 13L),
    axis.title.x = element_text(size = 13L))
```

### Level two categorical variable (Teaching Method)
```{r}
#Since the level 2 variables take the same value for each row of each classroom, keep the first row associated with each classroom
TestScoresLev2 <- 
  TestScores %>%
  group_by(classroom) %>%
  filter(row_number() == 1) 

#Find the count of classroom for each teaching method
TM_count <- table(TestScoresLev2$teaching_method)
kable(TM_count, format = "markdown", col.names = c("Teaching Method", "Count"), align = c("c", "c"))

#Find the proportion of classroom for each teaching method
TM_prop <- prop.table(table(TestScoresLev2$teaching_method))
kable(TM_prop, format = "markdown", col.names = c("Teaching Method", "Proportion"), align = c("c", "c"))

#Use bar graph to compare the number of classrooms for each teaching method
ggplot(data = TestScoresLev2, mapping = aes(x = teaching_method)) + 
  geom_bar(fill = "white", color = "black", width = 0.5) +
  labs(x = "Teaching Method",
       y = "Frequency",
       title = "Number of Classroom For Each Teaching Method") +
  theme(
    plot.title = element_text(size = 15L,
    face = "italic",
    hjust = 0.5), 
    axis.text.y = element_text(size = 11L),
    axis.text.x = element_text(size = 11L),
    axis.title.y = element_text(size = 13L),
    axis.title.x = element_text(size = 13L))
```


## EDA - bivariate
```{r}
#Use boxplot to visualize the distribution of posttest scores for different lunch type
ggplot(data = TestScores, mapping = aes(y = lunch, x = posttest)) +
  geom_boxplot() +
  labs(y = "Lunch Type",
       x = "Posttest Scores",
       title = "Boxplot of Posttest Score By Lunch Type") + 
  scale_y_discrete(labels = c("Does not qualify", "Qualifies for reduced/\nfree lunch")) +
  theme(
    plot.title = element_text(size = 15L,
    face = "italic",
    hjust = 0.5), 
    axis.text.y = element_text(size = 11L),
    axis.text.x = element_text(size = 11L),
    axis.title.y = element_text(size = 13L),
    axis.title.x = element_text(size = 13L))
```

```{r}
# Lattice plot for Posttest Scores vs. Lunch Type
ggplot(TestScores,aes(x = factor(lunch),y = posttest)) +  
  geom_dotplot(binaxis = "y",stackdir = "center",binwidth = 25/30) + 
  facet_wrap(~classroom,ncol = 7) + 
  theme(strip.text.x = element_blank()) + coord_flip() +
  labs(x = "Lunch Type",y = "Posttest Score") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10, margin = margin(r = 10)),
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5)
  )
```

```{r}
#Use side-by-side boxplots to explore the relationship between the posttest score and the teaching method by treating the observations as independent
posttest_TM <-
  ggplot(data = TestScores, mapping = aes(x = posttest, y = teaching_method)) +
  geom_boxplot() +
  labs(x = "Posttest Scores",
       y = "Teaching Method",
       title = "Boxplot of Posttest Score By Teaching Method") + 
  theme(
    plot.title = element_text(size = 15L,
    face = "italic",
    hjust = 0.5), 
    axis.text.y = element_text(size = 11L),
    axis.text.x = element_text(size = 11L),
    axis.title.y = element_text(size = 13L),
    axis.title.x = element_text(size = 13L))
```


```{r, warning=F, message=F}
#Create side-by-side boxplots to explore the relationship between the posttest score and the teaching method by finding the mean for each classroom type to aggregate the posttest score
posttest_mean_TM <-
  TestScores %>%
  group_by(classroom) %>%
  summarize(mean_posttest = mean(posttest, na.rm = T),
            teaching_method = teaching_method) %>%
  ggplot(mapping = aes(x = mean_posttest, y = teaching_method)) +
  geom_boxplot() +
  labs(x = "Mean Posttest Score",
       y = "Teaching Method",
       title = "Boxplot of Mean Posttest Score By Teaching Method \nFor Classrooms") +
  theme(
    plot.title = element_text(size = 15L,
    face = "italic",
    hjust = 0.5), 
    axis.text.y = element_text(size = 11L),
    axis.text.x = element_text(size = 11L),
    axis.title.y = element_text(size = 13L),
    axis.title.x = element_text(size = 13L))

grid.arrange(posttest_TM, posttest_mean_TM, ncol = 1)
```

## Modeling
```{r}
#Create encoding for each categorical variables. For level one variable lunch, 1=qualifies for reduced/free lunch, 0=does not qualify. For level two variable teaching_method, 1=standard, 0=experimental.
TestScores <-
  TestScores %>%
  mutate(qualify = ifelse(lunch == "Qualifies for reduced/free lunch", 1, 0),
         standard = ifelse(teaching_method == "Standard", 1, 0))
```


### Model with no covariates (random intercepts model)

$\textbf{Multilevel Model:}$ 

$$\text{Level 1: } Y_{ij} = a_i + \epsilon_{ij} \text{ where } \epsilon_{ij} \sim N(0, \sigma^2) $$

$$\text{Level 2: } a_i = \alpha_0 + u_i \text{ where } u_{i} \sim N(0, \sigma_u^2)$$

$\textbf{Composite Model:}$

$$Y_{ij} = \alpha_0 + u_i + \epsilon_{ij}$$

```{r}
model1 <- lmer(posttest ~ 1 + (1 | classroom), REML = FALSE, data = TestScores)
summary(model1)
```

$\hat\alpha_0 = 68.410$ This represents the average posttest score across all students and all classrooms. 

$\hat\sigma^2 = 10.38$ This is the $\epsilon_{ij}$ variance. It represents the estimated variance of within classroom deviations. 

$\hat\sigma^2_u = 183.35$ This is the $u_i$ variance. It represents the estimated variance of between classroom deviations. 

$$\hat{p} = \frac{\hat\sigma^2_u}{\hat\sigma^2_u+\hat\sigma^2} = \frac{183.35}{193.73} = 0.946$$

94.6% of the total variation in posttest scores is due to differences between classrooms. 

0.946 is the average correlation for any pair of students within the same classroom. 

### Level One Model

$\textbf{Multilevel Model:}$

$$\begin{align}
&\text{Level 1: } Y_{ij} = a_i + b_i x_{ij,\text{qualify}} + \epsilon_{ij}, \quad \text{where } \epsilon_{ij} \sim N(0, \sigma^2) \\
&\begin{aligned} \text{Level 2: }
a_i &= \alpha_0 + u_i \\
b_i &= \beta_0 + v_i
\end{aligned} \\
&\text{where } 
\begin{bmatrix}
u_i \\
v_i
\end{bmatrix}
\sim N\left(
\begin{bmatrix}
0 \\
0
\end{bmatrix}, 
\begin{bmatrix}
\sigma_u^2 & \sigma_{uv} = \rho_{uv} \sigma_u \sigma_v \\
\sigma_{uv} = \rho_{uv} \sigma_u \sigma_v & \sigma_v^2
\end{bmatrix}
\right)
\end{align}$$

$\textbf{Composite Model:}$

$$Y_{ij} = \alpha_0 + u_i + \beta_0x_{ij,qualify} + v_ix_{ij,qualify} + \epsilon_{ij}$$



```{r}
model2 <- lmer(posttest ~  qualify + (1 + qualify | classroom), REML = FALSE, data = TestScores)
summary(model2)
```

$|t| > 2$

We expect posttest scores to decrease by 3.39, on average, for students that qualify for free or subsidized lunches as compared to students that do not qualify. 

### Level Two Models

$\textbf{Multilevel Model:}$

$$\begin{align}
&\text{Level 1: } Y_{ij} = a_i + b_ix_\text{ij, qualify} + \epsilon_{ij}, \quad \text{where } \epsilon_{ij} \sim N(0, \sigma^2) \\
&\begin{aligned} \text{Level 2: }
a_i &= \alpha_0 + \alpha_1x_\text{i, standard} + u_i  \\
b_i &= \beta_0 + v_i
\end{aligned} \\
&\text{where } 
\begin{bmatrix}
u_i \\
v_i
\end{bmatrix}
\sim N\left(
\begin{bmatrix}
0 \\
0
\end{bmatrix}, 
\begin{bmatrix}
\sigma_u^2 & \sigma_{uv} = \rho_{uv} \sigma_u \sigma_v \\
\sigma_{uv} = \rho_{uv} \sigma_u \sigma_v & \sigma_v^2
\end{bmatrix}
\right)
\end{align}$$

$\textbf{Composite Model:}$

$$Y_{ij} = \alpha_0 + \alpha_1x_\text{i, standard} + \beta_0x_\text{ij, qualify} + u_i + v_ix_\text{ij, qualify} + \epsilon_{ij}$$

```{r}
#Random Intercept Model with Lunch and Teaching Method
model3 <- lmer(posttest ~ standard + qualify + (1 + qualify | classroom), REML = FALSE, data = TestScores)
summary(model3)
```


$\textbf{Multilevel Model:}$

$$\begin{align}
&\text{Level 1: } Y_{ij} = a_i + b_ix_{ij,qualify} + \epsilon_{ij}, \quad \text{where } \epsilon_{ij} \sim N(0, \sigma^2) \\
&\begin{aligned} \text{Level 2: }
a_i &= \alpha_0 + \alpha_1x_{i,standard} + u_i \\
b_i &= \beta_0 + \beta_1x_{i,standard} + v_i
\end{aligned} \\
&\text{where } 
\begin{bmatrix}
u_i \\
v_i
\end{bmatrix}
\sim N\left(
\begin{bmatrix}
0 \\
0
\end{bmatrix}, 
\begin{bmatrix}
\sigma_u^2 & \sigma_{uv} = \rho_{uv} \sigma_u \sigma_v \\
\sigma_{uv} = \rho_{uv} \sigma_u \sigma_v & \sigma_v^2
\end{bmatrix}
\right)
\end{align}$$

$\textbf{Composite Model:}$

$$Y_{ij} = \alpha_0 + \alpha_1x_{i,standard} + u_i + \beta_0x_{ij,qualify} + \beta_1x_{i,standard}x_{ij,qualify} + v_ix_{ij,qualify} + \epsilon_{ij}$$

```{r}
model4 <- lmer(posttest ~  qualify + standard + qualify:standard + (1 + qualify | classroom), REML = FALSE, data = TestScores)
summary(model4)
```

```{r}
anova(model1, model2, test = "Chisq")
anova(model1, model3, test = "Chisq")
anova(model1, model4, test = "Chisq")
anova(model2, model3, test = "Chisq")
anova(model3, model4, text = "Chisq")
```
